Closing the gap between statistical and scientific workflows for improved inference in ecology

### 2025-07-24 revision/response_r1.tex

\begin{mybox}
\emph{5. Theoretical examples provided, but no data application of the proposed workflow\\
Sections 3.1 and 3.2 and the Box outline theoretical examples of how the workflow could be applied for trend estimation and forecasting in the context of types of data, but there are no concrete data examples. A comment on how the proposed workflow could be applied to real data and how it would scale with complexity of data and research questions would be a good addition to the paper if it is not feasible to include real-data examples with the general discussion.}
\end{mybox}

% Real-data examples... not do-able!
...

\begin{mybox}
\emph{6. Adopting the workflow\\
Recommendations are made to train ecologists more quantitatively, but there is no suggestion or encouragement of collaboration with statisticians. These types of workflows are generally part of statistical practice, so could an alternative to more quantitative training (which would have to come at the cost of some other aspect of training) be more cross-disciplinary collaboration with statisticians?}
\end{mybox}

Good point. We have adjusted our language in one place (\href{file:forecastflows_r1\#lintarget:collab1}{line \ref*{lin:collab1}}) and we have added this suggestion also (\href{file:forecastflows_r1\#lintarget:collab}{line \ref*{lin:collab}}).

### 2025-07-24 revision/forecastflows_r1.tex
Edits by Lizzie finally

% \title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology} 
\title{Closing the gap between statistical and scientific workflows for improved inference in ecology} 
\date{\today}
\author{Victor Van der Meersch$^{1*}$, James Regetz$^{2}$, T. Jonathan Davies$^{1,3}$ \& EM Wolkovich$^{1}$}
\maketitle
\noindent $^{1}$ Forest and Conservation Sciences, University of British Columbia, Vancouver, BC V6T 1Z4, Canada\\
$^{2}$ National Center for Ecological Analysis and Synthesis, 1021 Anacapa St, Santa Barbara, CA 93101, United States\\
$^{3}$ Botany, University of British Columbia, Vancouver, BC V6T 1Z4, Canada \\
$^{*}$ \url{mailto:victor.vandermeersch@ubc.ca}\\

@@

\begin{abstract}
% Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare most current models for several reasons. First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models. Datasets relevant for developing these models have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models for several reasons. First, the research community is split between two disconnected paradigms: using data to fit simple, trend-like models with few parameters, or developing forecasting models that include complex, mechanistic submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform ecological modeling.

@@

Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets, and another focused on forecasting from generally distinct datasets or mechanistic models based on less data. 
Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced and suffer from large geographic, temporal and taxonomic biases. Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting.
Instead, forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning. These approaches have failed to yield clear agreement on current species trends, leading to ongoing debates about the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and producing forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012, Thuiller2019}.

\llabel{inferentialgoal}At first glance, these two modeling paradigms may appear to pursue independent inferential goals--- explanation vs.\ prediction---but in reality this distinction is blurred. Trend estimation may be nominally focused on explanatory outcomes, but especially because it tends to be carried out using simpler models, it often lacks a clear link with theories that facilitate meaningful causal interpretation. Moreover, whether implicitly or explicitly, estimated trends are frequently treated as quantities that project forward into the future, taking on a forecasting role. Meanwhile, although more complex forecasting models are certainly built for predictive purposes, the typical use of mechanistic, process-based model components is often justified by the assumption that ``we must understand to predict”---and indeed, such models are often judged by how faithfully they mimic the real-world processes they represent---which contributes to the conflation of explanation with prediction \citep{Shmueli2010}.

We argue that current debates, confusions and diverging forecasts are driven in large part by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Research estimating biodiversity trends has become focused on methodological aspects; the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting. At the same time, process-based models developed for forecasting often evolve through the addition of new separate layers or components. These new parts are often disconnected from the original research aim, its data stream, and the previous scientific insights, because current approaches rarely examine the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below).

@@

\begin{figure}
	\centering
	\includegraphics{../figures/figure_modeldataspaces_revised}
	\caption{Different trajectories through model-data space during model development, from pre-model/pre-data to post-model/post-data quadrants.
	Panel A serves as a legend, showing where model development should begin (carefully thinking about the hypotheses we want to test) and when inference should be made, for either explanatory purpose (estimation) or forecasting (prediction).
	Panel B illustrates that in current trend estimation approaches, most time is spent in the post-model/post-data quadrant, where models are directly fitted to empirical data---often with no prior evaluation of the model.
	Panel C illustrates the typical process-based model development. A lot of time is spent building and calibrating (fitting) separate submodels (represented by the separated ovals), without clear distinction between model building and model calibration and without evaluating the full model structure a priori. Additionally, the specific hypotheses the model is designed to test are rarely clearly stated.
	Panel D illustrates the workflow highlighted in this paper (see main text and Figure 2 for details). Ideally, it begins in the pre-model/pre-data quadrant, where the research question is defined. The next step moves to the post-model/pre-data quadrant, where a model is built---independently of the data. This is where we should spend a good deal of time evaluating model behavior using simulated data. Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data. Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedback to refine our model.}
	\label{fig:modeldata}
\end{figure}

@@

This reality should drive researchers to use more robust and coherent methods. But the current workflows combined with the challenges ecologists are facing---both in terms of data complexity and societal needs---instead may lead to persistent problems.
Trend estimation has focused mostly on fitting a model to empirical data (\llabel{quad1}i.e. post-model/post-data, figure \ref{fig:modeldata}b)---without the checks (\llabel{quad2}post-model/pre-data) and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (figure \ref{fig:modeldata}b). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models (figure \ref{fig:modeldata}c), frequently obscuring the steps underlying model building and parameterization (\llabel{quad3}and without a clear distinction between pre- and post-data). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.

We argue a workflow that moves \llabel{quad4}along the data-model space in a coherent sequence of steps (figure \ref{fig:modeldata}d) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, figure \ref{fig:workflow}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error (\llabel{quad5}pre-model/pre-data, figure \ref{fig:modeldata}d).  
\llabel{causalinf}This step of carefully formulating hypotheses has gained increasing attention in the ecological literature \citep{Grace2020}. From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, figure \ref{fig:workflow}). The general idea is to start with a relatively simple model that we could refine later (see example workflow we provide). At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterizations).

With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, figure \ref{fig:workflow}). `Fake' or `test' data are generated directly from the model by fixing parameters \llabel{firstsim}to some reasonable values (which is straightforward if the parameters are interpretable) and from fake predictor data. 
We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values (\llabel{quad6}post-model/pre-data, figure \ref{fig:modeldata}d). At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is realistic or not. This is also a step for making sure the model is working as expected (see example worklow). \llabel{repeatsim}Ideally, this data simulation step should be repeated several times with different parameter values, sampled within a reasonable range. \llabel{priorPC}In a Bayesian framework, simulated parameters can be sampled from the prior distribution (allowing us to check the implications of our priors, \citealp{Gelman2020}).

\begin{figure}[h]
	\centering
	\vspace*{-0.5cm}
    \hspace*{-1.5cm}
	\includegraphics{../figures/figure_worflow_wsteps}
	\caption{The workflow we propose here \citep[which builds from recent advances in workflows,][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024} focuses on iterative feedbacks between the research question, model and data. Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5). Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.  The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream. This figure is expanded from Figure \ref{fig:modeldata}d.}
	\vspace*{-0cm}
	\label{fig:workflow}
\end{figure}

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, figure \ref{fig:workflow}). This way, we obtain parameter estimates constrained by observations (\llabel{quad7}post-model/post-data, figure \ref{fig:modeldata}d). 
These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, figure \ref{fig:workflow}). This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations. \llabel{howtoPPC}There is no general way to do this: the goal is to tailor checks that assess key features of the data in regards to our research question \citep{Gelman2020}.
It's only once all steps have been completed that we can interpret parameter values with respect to our research question.
The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting.

Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data. The adoption of the workflow in macroecological studies---where model building is often informed by patterns in the data---would also make the exploratory analysis more transparent (as an explicit preliminary step for entering the workflow) and would compel researchers to more clearly develop a research question before extensive model fitting.

A key feature of this workflow is the central role of data simulation, which introduces two feedback loops---\llabel{broadcontext}and should be applied not only in a Bayesian framework but across broader modeling approaches.
The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated),
flagging the need to change the model structure---before incorporating empirical observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es) (if we can identify them) and return to the start of the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased, this bias should be built into the model and thus into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.

@@

Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to accurately \llabel{quant1}understand and manage uncertainties and yield new scientific insights. This will help refocus the debate on designing new hypotheses, formulating new questions---and guiding efforts to collect new data. Here, we illustrate how such a workflow could lead to significant improvements in two case studies: (i) estimating global biodiversity trends and (ii) forecasting future species and ecosystem dynamics using process-based models.

@@

An improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.  Using the workflow would make it more obvious that what now appear as major discrepancies are shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process. Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected. But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model. We expect retrodictive checks on most published trend analyses would highlight major missing components in these models, and drive changes both in the models themselves and in the simulated data to check the models. This step builds somewhat on the skills needed for null models \citep{Gotelli:2012oz}, but with a shift in focus towards the specific bespoke model at hand. Ecologists have started to use simulated data more to understand potential limitations of their models and data combined \citep{Hilborn1997}, but this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility if used in an iterative workflow.

Applying the workflow to current trend estimates could importantly highlight the best way to improve data collection for more reliable estimates. Returning to the example of a global estimate of trends in vertebrate populations of species over time (see Box) and applying our proposed workflow would mean more efforts to define the goal and question---is it a simple global estimate?
Or a need to also find which species are declining most, including those that may have poor or no data? From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or 'synthetic data generation' currently \citep{Buschke2021,mcrae2025utility} but could be built into the models fit to the empirical data. Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model and, ultimately, this would help inform our global estimates of mean trends (see workflow example). For example, certain populations are recovering for very specific reasons (e.g., elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled. From this model, what data are most critically needed to address the updated aims would become clearer and could drive new data collection \citep{toszogyova2024mathematical}.

@@

\centerline{\bf Trend estimation}
\vspace*{2mm}
\begin{minipage}[t]{\linewidth}
    \includegraphics[width=\linewidth]{../figures/trendestimation_details}

@@

\subsection{[Need a better title] A tractable alternative to machine learning}

With rapid advances in machine learning, improving current methods to gain greater scientific insights to drive better forecasting seems increasingly important. \llabel{MLvsPBM}Machine learning will likely surpass process-based models for forecasting accuracy if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability. Similarly, estimates of trends from empirical data using models without clear mechanistic drivers may soon offer fewer advantages over machine learning.

>>>>

% \title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology} 
\title{Closing the gap between statistical and scientific workflows for improved inference in ecology} %emw24Jul: Hmm, you say in the letter you changed the title but I am not sure you did. 
\date{\today}
\author{Victor Van der Meersch$^{1*}$, James Regetz$^{2}$, T. Jonathan Davies$^{1,3}$ \& EM Wolkovich$^{1}$}
\maketitle
\noindent $^{1}$ Forest and Conservation Sciences, University of British Columbia, Vancouver, BC V6T 1Z4, Canada\\
$^{2}$ National Center for Ecological Analysis and Synthesis, 1021 Anacapa St, Santa Barbara, CA 93101, United States\\
$^{3}$ Botany, University of British Columbia, Vancouver, BC V6T 1Z4, Canada \\
$^{*}$ \url{mailto:victor.vandermeersch@ubc.ca}\\

@@

\begin{abstract}
% Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare most current models for several reasons. First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models. Datasets relevant for developing these models have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. 
%emw24Jul: Are you sure you want 'paradigm'? I love the word and the sentence but I think we mean more two 'groups' or 'spheres' -- I would switch and then carry the language you use to the next sentence. 
We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models for several reasons. First, the research community is split between two disconnected paradigms that prevent uniting a natural connection between trend estimation and forecasting. One research sphere focuses on using data to fit simple, trend-like models with few parameters, which is separated from research developing forecasting models, which often include complex, mechanistic sub-models only indirectly informed by data. Second, in both cases, models tend to be developed without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions throughout the entire modeling process. We propose a principled workflow that can address both of these limitations, by integrating statistical and scientific practices through clear steps, many of which rely on data simulation to inform decisions in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could improve our understanding of ecological systems and forecasts for them.

%emw24Jul: I pulled out a few of the best sentences that may help the abstract: 
% Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date, alongside the skill to forecast future dynamics. 
% Towards this aim, we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.
% Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.

@@

Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets, and another focused on forecasting from generally distinct datasets or mechanistic models based on fewer data. 
Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced and suffer from large geographic, temporal and taxonomic biases. Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting.
Instead, forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning. These approaches have failed to yield agreement on current species trends, leading to ongoing debates about the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and producing forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012, Thuiller2019}.

%emw24Jul: Same comment as above about 'paradigms' but here I think you mean 'approaches' so I just changed it. 
\llabel{inferentialgoal}These two modeling approaches may appear to pursue independent inferential goals---explanation vs.\ prediction---but in reality this distinction is blurred. Trend estimation may be nominally focused on explanatory outcomes, yet it rarely links clearly to theories that could provide causal inference, usually because it tends to be carried out using simpler models. Moreover,  estimated trends are frequently treated---implicitly and explicitly---as quantities that project forward into the future, taking on a forecasting role. %emw24Jul: Nice!
In contrast, approaches using complex forecasting models are certainly built for predictive purposes, yet these also have an additional agenda. Developing and using mechanistic, process-based models is often justified by the assumption that ``we must understand to predict," and such models are often judged by how faithfully they mimic the real-world processes they represent, which conflates explanation with prediction \citep{Shmueli2010}.

%emw24Jul: I edited something out that I do and you should avoid (or try to edit out when I add it): adding too many 'subjunctive/conditional' phrases. We ARGUE THAT MANY. ... IN LARGE PART mean ... try to pick one (of two here) conditional phrase and drop the rest. 
We argue that many current debates and diverging forecasts are driven by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Research estimating biodiversity trends has become focused on methodological aspects because the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting. At the same time, process-based models developed for forecasting often evolve through the addition of new separate layers or components because the current workflow rarely examines the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below). Thus it encourages new parts that are often disconnected from the original research aim, its data stream, and the previous scientific insights.

@@

\begin{figure}
	\centering
	\includegraphics{../figures/figure_modeldataspaces_revised}
	\caption{Different trajectories of current (b-c) and proposed (d) workflows through model-data space during model development, from pre-model/pre-data to post-model/post-data quadrants.
	Panel A serves as a legend, showing where model development should begin (carefully thinking about the hypotheses to test) and when inference should be made, for either explanatory purpose (estimation) or forecasting (prediction).
	Panel B illustrates that, in current trend estimation approaches, most time is spent in the post-model/post-data quadrant, where models are directly fitted to empirical data---often with no prior evaluation of the model.
	Panel C illustrates the typical process-based model development where a lot of time is spent building and calibrating (fitting) separate submodels (represented by the separated ovals) without clear distinction between model building and model calibration and without evaluating the full model structure a priori. Additionally, the specific hypotheses the model is designed to test are rarely clearly stated.
	Panel D illustrates the workflow highlighted in this paper (see main text and Figure 2 for details). Ideally, it begins in the pre-model/pre-data quadrant, where the research question is defined. The next step moves to the post-model/pre-data quadrant, where a model is built---independently of the data. This is where we should spend a good deal of time evaluating model behavior using simulated data. Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data. Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedback to refine our model.}
	\label{fig:modeldata}
\end{figure}

@@

%emw24Jul: Nice stuff here!
This reality should drive researchers to use more robust and coherent methods. But the current workflows combined with the challenges ecologists are facing---both in terms of data complexity and societal needs---instead may lead to persistent problems.
Trend estimation has focused mostly on fitting a model to empirical data (\llabel{quad1}i.e. post-model/post-data, figure \ref{fig:modeldata}b)---without the checks (\llabel{quad2}post-model/pre-data) and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (figure \ref{fig:modeldata}b). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models (figure \ref{fig:modeldata}c), frequently obscuring the steps underlying model building and parameterization (\llabel{quad3}and without a clear distinction between pre- and post-data). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.

We argue a workflow that moves \llabel{quad4}along the data-model space in a coherent sequence of steps (figure \ref{fig:modeldata}d) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, figure \ref{fig:workflow}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error (\llabel{quad5}pre-model/pre-data, figure \ref{fig:modeldata}d).  
\llabel{causalinf}This step of carefully formulating hypotheses has gained increasing attention in the ecological literature \citep{Grace2020}. From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, figure \ref{fig:workflow}). Generally, starting with a relatively simple model that we could refine later makes understanding the model and how it interacts with empirical data easier (see example workflow we provide). At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterizations). 
%emw24Jul: 'see example workflow we provide' and 'see example workflow' -- can we standardize our mention of this and make it clear where it is? Like: `see example workflow we provide as supplement files.'

With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, figure \ref{fig:workflow}). `Fake' or `test' data are generated directly from the model by fixing parameters \llabel{firstsim}to some reasonable values, which is straightforward if the parameters are interpretable, and from `fake' (simulated) predictor data. 
We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values (\llabel{quad6}post-model/pre-data, figure \ref{fig:modeldata}d). At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is realistic or not. This is also a step for making sure the model is working as expected (see example worklow). \llabel{repeatsim}Ideally, this data simulation step should be repeated several times with different parameter values, sampled within a reasonable range. \llabel{priorPC}In a Bayesian framework, simulated parameters can be sampled from the prior distribution (allowing us to check the implications of our priors \citealp{Gelman2020}), while in other frameworks they can be sampled from a range of biologically reasonable values (which is very similar to the concept of a prior).

\begin{figure}[h]
	\centering
	\vspace*{-0.5cm}
    \hspace*{-1.5cm}
	\includegraphics{../figures/figure_worflow_wsteps}
	\caption{The workflow we propose here, which builds from recent advances in workflows \citep[][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024}, focuses on iterative feedbacks between the research question, model and data. Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5). Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.  The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream. This figure is expanded from Figure \ref{fig:modeldata}d.}
	\vspace*{-0cm}
	\label{fig:workflow}
\end{figure}

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, figure \ref{fig:workflow}). This way, we obtain parameter estimates constrained by observations (\llabel{quad7}post-model/post-data, figure \ref{fig:modeldata}d). 
These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, figure \ref{fig:workflow}). This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations. \llabel{howtoPPC}This is a powerful approach, in part because checks should be tailored to assess key features of the data and model output together in regards to our research question, but this also means there is no standard way to do this \citep{Gelman2020}.
It is only once all steps have been completed that we can interpret parameter values with respect to our research question.
The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting.

Within such a workflow, forecasting emerges as a natural outcome. Rather than being a final goal, forecasting only involves jointly modeling new circumstances along with the original data. This workflow could also bridge gaps in certain areas between exploratory analyses and developing research questions. For example, in macroecological studies model building is often informed by patterns in the data. The adoption of this workflow would make the exploratory stage more transparent---as an explicit preliminary step for entering the workflow---and would compel researchers to more clearly develop a research question before extensive model fitting. %emw24Jul: possible to ref Fig 1 here?

A key feature of this workflow is the central role of data simulation, which introduces two feedback loops.
The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated),
flagging the need to change the model structure---before incorporating empirical observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es)---if we can identify them---and return to the start of the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased, this bias should be built into the model and thus into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes. \llabel{broadcontext}These feedback loops are applied currently in only in a Bayesian frameworks (to our knowledge), but could and should extend across other modeling approaches given their power to improve models and, in turn, scientific understanding.

@@

%emw24Jul: I removed case study from this draft, I think they're more than case studies and the term no longer helps the paper.
Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to accurately \llabel{quant1}understand and manage uncertainties and yield new scientific insights. This will help refocus the debate on designing new hypotheses, formulating new questions---and guiding efforts to collect new data. Here, we illustrate how such a workflow could lead to significant improvements in two related areas: (i) estimating global biodiversity trends and (ii) forecasting species and ecosystem dynamics using process-based models.

@@

%emw24Jul: Cite supp retro check figure here below? 
We argue that an improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.  Using the workflow would make it more obvious that what now appear as major discrepancies are shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process. Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected. But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model.

We expect retrodictive checks on most published trend analyses would highlight major missing components in these models, and drive changes both in the models themselves and in the simulated data to check the models. This step builds somewhat on the skills needed for null models \citep{Gotelli:2012oz}, but with a shift in focus towards the specific bespoke model at hand. Ecologists have started to use simulated data more to understand potential limitations of their models and data combined \citep{Hilborn1997}, but this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility if used in an iterative workflow.

Applying the workflow to current trend estimates could importantly highlight the best way to improve data collection for more reliable estimates. Returning to the example of a global estimate of trends in vertebrate populations of species over time (see Box) and applying our proposed workflow would mean more efforts to define the goal and question---is it a simple global estimate?
Or a need to also find which species are declining most, including those that may have poor or no data? From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or `synthetic data generation' currently \citep{Buschke2021,mcrae2025utility}, but could be built into the models fit to the empirical data. Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model and, ultimately, this would help inform our global estimates of mean trends (see workflow example). For example, certain populations are recovering for very specific reasons (e.g., elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled. From this model, what data are most critically needed to address the updated aims would become clearer and could drive new data collection \citep{toszogyova2024mathematical}.

@@

\centerline{\bf Trend estimation} %emw24Jul: I did not have time to stare at the figures to make sure they make sense ... if there are any that need a second set of eyes, let me know and I can. 
\vspace*{2mm}
\begin{minipage}[t]{\linewidth}
    \includegraphics[width=\linewidth]{../figures/trendestimation_details}

@@

%emw24Jul: Bold proposed changes below.... see what you think. 
% \subsection{[Need a better title] A tractable alternative to machine learning}
\emph{Conclusions:}

With rapid advances in machine learning, improving current methods to gain greater scientific insights to drive better forecasting seems increasingly important. \llabel{MLvsPBM}Machine learning will likely surpass process-based models for forecasting accuracy, especially if the latter lack a robust estimation of their parameters and fall in a complexity trap---a trap with additional costs to interpretability of process-based models. Similarly, estimates of trends from empirical data using models without clear mechanistic drivers may soon offer fewer advantages over machine learning.
